### 2016-11-24 chapter1.tex
Corrections made to Ionosphere Section

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The Earth's ionosphere is a region of the upper atmosphere that ranges approximately between 50-1000 km in altitude where neutral gases have been excited so that they are partially ionoized, resulting in free electrons and ions mixed with neutral particles.  Two conflicting processes occur in the ionosphere to create a peak in electron density, commonly referred to as the Chapman Layer \textbf{CTIE CHAPMAN?}.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increase as altitude decreases because there are more neturals available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination into neutral particles becomes a substantal factor, reducing the ion density.  The exact altitude where this density peak occurs is variable with time of day, location, season, and solar cycle.

@@

\subsection{Ionosphere Regions}
\label{sec:ionosphere_regions}
The ionosphere has historically been divided into three distinct region.  The D region is located between 80 and 90 km in altitude and typically only exists in the daytime when the ionosphere is sunlit.  The E region typically exists between 90 and 120 km with a density peak between 105 and 110 km, depending on factors such as time of day and season.  Above 120 km is considered the F region, which has a density peak between 300 and 400 km.  The F region can actually have two peaks under some daytime conditions, notated as the F1 peak and the F2 peak.

Because the concentrations of ions and neutral particles changes with altitude, each of these regions have distinct plasma physical properties that make them unique and change how plasma structuring forms.  In the E region, electrons are magnetized so their motion is mostly due to gyration around the magnetic field.  However, the motion on ions is dominated by collisions between ions and neutrals, so ions in the E region are considered collisional.  In the F region where the neutral density is much lower, both ions and electrons are magnetized.  These differences vary the conductivity in both regions and can strongly impact how plasma waves develop, which will be discussed further in section \ref{sec:lit_experiment}.

\subsection{Plasma Structuring in the Polar Ionosphere}
\label{sec:polar_structure}
The polar ionosphere is a particularly interesting and dynamic region due to the presence of the magnetic poles and open field lines to the solar wind.  Plasma structuring can occur as variations in density, velocity, or temperature, but the main focus of this work will be density perturbations, often called irregularities.  Density irregularities in the polar cap can occur on scales ranging from thousands of kilometers to less than a centimeter.  Large-scale plasma irregularities are typically considered to be any density structuring on the scale of 1 -- 1000 km.  Some common examples are polar patches, polar holes, and sun-aligned arcs.  Polar patches are large density enhancements (usually at least twice the background plasma density) that travel across the polar cap with the background convection.  Polar holes are plasma density depletions that typically occur in the F region slightly poleward of the auroral oval.  Sun-aligned arcs are large density forms that stretch across the polar cap along the sun-earth line.  They are very narrow and often characterized by large and complex velocity shears on either side of the arc.

Intermediate-scale structuring typically ranges from 100 m -- 1 km.  This is also referred to as scintillation-causing structuring and is responsible for many of the negative space weather effects that are observed on Earth.  Radio signal scintillation through the ionosphere changes the phase and amplitude of the original signal, which can effect the ability that satelites have to communicate with ground recievers.  This can cause communication blackouts or introduce large errors into GPS calculations, effecting navigation.  Because communication and navigation technology is implemented in both every day civilian life as well as government and military intrests, the effects of this can be far reaching.

Small-scale structuring is generally considered to be all irregularities less than 100 m.  Although the effects of these scale sizes on navigation and communication systems are minimal, they can be the easiest to study because of large data sets collected from HF radars globally for several decades.  The irregularity sizes presented here are rough categories only and a high degree of coupling between different scales is generally considered reasonable.  For instance, large-scale structures such as polar patches may very well have internal intermediate- or small-scale structuring.  Some kind of turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs.

@@

\subsection{Coherent Scatter Radars}
\label{sec:csr}

Coherent Scatter Radars (CSR) have been used to study the Ionosphere for the last half century.  Radars are ground based instruments used to detect density structures in the ionosphere's plasma.  A basic radar consists of transmission and recieveing antenna.  The radar transmits a radio wave, which can reflect off a suititbly sized structure in the ionosphere.  Reflected waves can be detected by the recieving antennas.  The time difference between when the signal was transmitted and when it was detected  an then be used to determine how far away from the radar the target is.

There are three measurements typically made by CSR systems: backscatter power, line of sight (LoS) velocity, and spectral width.  Backscatter power is the power of the retuned signal that the radar's recievers detect.  Typically, a threshold is selected for the minimum power acceptable for a return to be considered an actual signal instead of background noise.  LoS velocity is the component of the total velocity that is measured along one beam of a radar.  Because radars measure velocity through the doppler shift of backscatter, only the component of the velocity vector that is along the radar beam can be measured.  Spectral width represents the width of the doppler power spectrum.

%- Greenwald, 1985

>>>>

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The Earth's ionosphere is a region of the upper atmosphere that ranges approximately between 50--1000 km in altitude where neutral gases have been excited so that they are partially ionoized, resulting in free electrons and ions mixed with neutral particles.  Two conflicting processes occur in the ionosphere to create a peak in electron density, commonly referred to as the Chapman Layer \textbf{CTIE CHAPMAN?}.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increase as altitude decreases because there are more neturals available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination into neutral particles becomes a substantal factor, reducing the ion density.  The exact altitude where this density peak occurs is variable with time of day, location, season, and solar cycle.  A profile of how electron density changes with altitude is shown in Figure \ref{fig:densprofile}.

@@

Above 50 km, the primary neutral particles are atomic oxygen, O, molecular oxygen, \(O_2\), and molecular nitrogen, \(N_2\).  Common ions at these altitudes are therefor mostly composed of oxygen and nitrogen, \(O^+, NO^+, O_2^+, N_2^+\), but hydrogen, \(H^+\), can contribute as well \citep{Kelley2009}.  In the ionosphere, neutral densties are typically several orders of magnitude higher than ion densities, however even a small number of charged particles changes the behavior of the region dramatically.  Because there are so many different ion species, plasma density is usually discussed simply in terms of electron density.

\subsection{Ionosphere Regions}
\label{sec:ionosphere_regions}
The ionosphere is typically divided into three region, Figure \ref{fig:densprofile}.  The D region is located between 80 and 90 km in altitude and typically only exists in the daytime when the ionosphere is sunlit.  The E region typically exists between 90 and 120 km with a density peak between 105 and 110 km, depending on factors such as time of day and season.  Above 120 km is considered the F region, which has a density peak between 300 and 400 km.  The F region can actually have two peaks under some daytime conditions, notated as the F1 peak and the F2 peak.  The locations and peak densities of each of these regions is highly variable both diurnally and seasonally.  In Figure \ref{fig:densprofile}, the daytime profile is shown with the solid line next to a nighttime profile with a dashed line.  Note that night electron densities are lower at all altitudes due to the lack of photoionization durring nighttimes.  In addition, the F-region peak moves to higher altitudes at night.

Because the concentrations of ions and neutral particles changes with altitude, each of these regions have distinct plasma physical properties that make them unique and change how plasma structuring forms.  The main change that occurs is wether the the large-scale motion of a particle is controlled more by the magnetic field (the particle is magnetized) or by collisions with other particles (collisional).  This can be determined by comparing the collision frequency, \(\nu_\alpha\), to the gyrofrequency, \(\Omega_\alpha\), of a particular species.  If \(\Omega_\alpha \gg \nu_\alpha\), the particle completes many gyrations between each collision, so its motion is determined mostly by the magnetic field and it is magnetized.  If \(\nu_\alpha \gg \Omega_\alpha\), the particle collides with other particles much more often than it complete a gyration, so it is considered collisional.  The gyrofrquency is dependent on the mass and charge of the particle and the strength of the magnetic field, \(\Omega_\alpha = q_\alpha B/m_\alpha\), so is roughly constant through the ionosphere because the magnetic field strength does not change much through these altitudes.  The collision frequency depends on the density of a species relative to the density of all other species in the plasma and can be calculated from a series of complicated expressions \citep{Schunk1980,Schunk2009}.  In the E region, \(\Omega_e \gg \nu_e\), so electrons are magnetized.  However, the motion on ions is dominated by collisions between ions and neutrals (\(\nu_i > \Omega_i\)), so ions in the E region are considered collisional.  In the F region where the neutral density is much lower, both ions and electrons are magnetized, \(\Omega_\alpha \gg \nu_\alpha\).  These differences in how charged particles move vary the conductivity and can strongly impact how plasma waves develop, which will be discussed further in Section \ref{sec:lit_theory}.

\subsection{Plasma Structuring in the Polar Ionosphere}
\label{sec:polar_structure}
The polar ionosphere is a particularly interesting and dynamic region due to the presence of the magnetic poles and open field lines to the solar wind.  Plasma structuring can occur as variations in density, velocity, or temperature, but the main focus of this work will be density perturbations, often called irregularities.  Density irregularities in the polar cap can occur on scales ranging from thousands of kilometers to less than a centimeter \citep{Tsunoda1988}.  Large-scale plasma irregularities are typically considered to be any density structuring on the scale of 1 -- 1000 km.  Some common examples are polar patches, polar holes, and sun-aligned arcs.  Polar patches are large density enhancements (usually at least twice the background plasma density) that travel across the polar cap with the background convection.  Polar holes are plasma density depletions that typically occur in the F region slightly poleward of the auroral oval.  Sun-aligned arcs are large density forms that stretch across the polar cap along the sun-earth line.  They are very narrow and often characterized by large and complex velocity shears on either side of the arc.

Intermediate-scale structuring typically ranges from 100 m -- 1 km.  This is also referred to as scintillation-causing structuring and is responsible for many of the negative space weather effects that are observed on Earth.  Radio signal scintillation through the ionosphere changes the phase and amplitude of the original signal, which can effect the ability that satelites have to communicate with ground recievers.  This can cause communication blackouts or introduce large errors into GPS calculations, effecting navigation.  Because communication and navigation technology is implemented in both every day civilian life as well as government and military intrests, the effects of this kind of signal scintillation can be far reaching.

Small-scale structuring is generally considered to be all irregularities less than 100 m.  Although the effects of these scale sizes on navigation and communication systems are minimal, they can be the easiest to study because of large data sets collected from high frequency ionospheric radars globally for several decades (discussed further in Section \ref{sec:superdarn}).  The irregularity sizes presented here are rough categories only and a high degree of coupling between different scales is generally considered reasonable.  For instance, large-scale structures such as polar patches may very well have internal intermediate- or small-scale structuring.  Some kind of turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs.

@@

\subsection{Coherent Scatter Radars}
\label{sec:csr}
Coherent Scatter Radars (CSR) have been used to study the ionosphere for the last half century.  The radar transmits a radio wave, which can reflect off plasma structuring in the ionosphere, and then be detected by the recieving antennas.  The time difference between when the signal was transmitted and when it was detected and the power and phase of the returned signal can then be used to determine local characteristics of ionospheric plasma.

There are three measurements typically made by CSR systems: backscatter power, line of sight (LoS) velocity, and spectral width.  Backscatter power is the power of the retuned signal that the radar's recievers detect.  Typically, a threshold is selected for the minimum power acceptable for a return to be considered an actual signal instead of background noise.  LoS velocity is the component of the total velocity that is measured along one beam of a radar.  Because radars measure velocity through the doppler shift of backscatter, only the component of the velocity vector that is along the radar beam can be measured.  Spectral width represents the width of the doppler power spectrum \citep{Greenwald1985}.
